---
title: "Writing a great story for data science projects - Fall 2025 "
subtitle: "This is a Report Template Quarto"
author: "Namita Mishra (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

<!-- -->

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

Example of writing including citing references:

*This is an introduction to ..... regression, which is a non-parametric
estimator that estimates the conditional expectation of two variables
which is random. The goal of a kernel regression is to discover the
non-linear relationship between two random variables. To discover the
non-linear relationship, kernel estimator or kernel smoothing is the
main method to estimate the curve for non-parametric statistics. In
kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]*. Topology can be used in machine
learning [@adams2021topology]

For Symbolic Regression [@wang2019symbolic] *This is my work and I want
to add more work...*

Cite new paper [@su2012linear]

#Introduction#

1. Forecasting the occurrence and non-occurrence of any disease is crucial in disease management. It requires long experimental procedures and non-outbreak prediction are not explored. Timely prediction of disease happenings could support effective implementation of preventative measures through early detection of disease outbreaks and non-outbreaks, providing valuable insights for mitigation strategies of novel diseases. [@gao2025early]

2. Rather than using expensive molecular information which is not easily available,the author identifies subjects at of cardiovascular disease (CVD),  thrugh the use of an alternative and an efficient approach to identifying by incorporating routinely-checked biological markers a cluster of multiple CVD risk factors like the metabolic syndrome, to classify the risk of underlying population by dint of Metabolic scores information  together with conventional risk factors (gender, heredity, smoking and other relevant factors). It is considered that different ethnic groups and areas with varied levels of socioeconomic status, have their own models which may not be applied to other areas with different socioeconomic status, because the baseline incidence (prior) varies from place to place. Such heterogeneity is often unobservable and unmeasurable. The physician may need to be wary of this baseline information before other clinical and individual variables are added. 
The study based to predict at risk individuals for CVD at individual level based on the information gleaned from the past.

The individual risk is elicited by prior speculation (first impression) which is then updated by objective observed data (patient's history and laboratory findings) the concept called is called Bayesian clinical reasoning [9]. It models the clinical weights (regression coefficients for computing risk score) as a random variable specified by certain statistical distribution (e.g. normal distribution) rather than a fixed value that is assumed in the traditional risk prediction model based on a frequentist viewpoint. Credible intervals of predicted risk estimates were obtained by averaging out the uncertainty of parameters (clinical weights) to reflect the precision of predicting individual risk. Bayesian approach update the prior distribution with the likelihood based on current data into posterior distribution, which then turns into tomorrow's prior distribution when new evidence is entered.

Case history: 
For a male at a clinic for routine health evaluation, his health status based on prior information of a 5-year CVD risk assessment is 11.2% (95% CI: 7.8%– 15.6%) which was not sufficient enough to alert him for any intervention needed. But, after history-taking,, the added new information of smoking with over 2 pack per-day and family history of CVD (given the likelihood ratios of 1.62 and 1.47, respectively,), the bayesian gave the updated 5-year CVD risk by adding this new information with an increased risk to 30.9% (95% CI: 20.7%–39.8%). In Bayesian, the prior probability of CVD via a logistic regression model is incorporated to the five-year longitudinal follow-up data on newly diagnosed incident CVD cases, in a sequential approach to predict the individual risk prediction for CVD. This accounts for having a universal predictive model without considering such an ecological influence. After the consideration of ecological heterogeneity, the individual-level risk prediction for the risk for CVD, urges the patient to take a life-modification program (quit smoking, healthy diet and regular exercise) [@Liu2013]


3. In Bayesian inference, all unknowns—observable (data) as well as unobservables (parameters and auxiliary variables)—are considered to be random and assigned probability distributions.These distributions best summarize the available information, and Bayesian inference then updates prior knowledge about the unobservables with information about them contained in the data.
The prior distribution and likelihood function are both necessary parts of a Bayesian analysis. Depending on the strength of information contained in the measurements, the prior distribution may or may not influence the outcome or parts thereof. [@Klauenberg2015]


4. Using a Bayesian approach, This study projected the increase in the rates of Atherosclerotic cardiovascular disease (ASCVD) in Asian countries. Given the significant personal and societal burden, the study aimed to accurately forecast ASCVD incidence rates to inform health care policy development and investments to targeted population.
Bayesian age-period-cohort (APC) models were then fitted to the observed data (1990-2021) to project age-standardized rates (with the mid-year 1990 population as the index) for the prevalence and mortality for IHD, stroke, and PAD from 2022 through 2040. Default log-normal priors chosen for each term in the model and the posterior estimates from the model fit were obtained to predict the incidence rates. [@Cotton2025]





## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}

# Load Data
kable(head(murders))


ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References


1.Gao S, Chakraborty AK, Greiner R, Lewis MA, Wang H. Early detection of disease outbreaks and non-outbreaks using incidence data: A framework using feature-based time series classification and machine learning. PLOS Computational Biology [Internet]. 2025 Feb 1 [cited 2025 Aug 30];21(2):e1012782. Available from: https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1012782

2. Liu, Y. M., Chen, S. L. S., Yen, A. M. F., & Chen, H. H. (2013). Individual risk prediction model for incident cardiovascular disease: A Bayesian clinical reasoning approach. International Journal of Cardiology, 167(5), 2008–2012. https://doi.org/10.1016/j.ijcard.2012.05.016

3. Klauenberg, K., Wübbeler, G., Mickan, B., Harris, P., & Elster, C. (2015). A tutorial on Bayesian Normal linear regression. Metrologia, 52(6), 878–892. https://doi.org/10.1088/0026-1394/52/6/878Links to an external site.

4. Cotton, A., Pedro RVO Salerno, M., Zhuo Chen, P., Salim Virani, M., Naveed Sattar, M., Sanjay Rajagopalan, M., & Salil V. Deo, M. (2025). Forecasting Atherosclerotic Cardiovascular Disease in South Asia Until 2040: A Bayesian Modeling Approach. JACC: Asia. https://doi.org/10.1016/J.JACASI.2025.06.014
